Linkage disequilibrium can drive strong and unstable variances and covariances in fitness-associated traits

### 2018-02-12 manuscript.tex
added new references, updated local copy of the manuscript, and changed figure 4 in the plots code.

\documentclass[9pt,twocolumn,twoside]{gsajnl}
% Use the documentclass option 'lineno' to view line numbers
\usepackage{caption}
\usepackage{subcaption}
\usepackage[nameinlink,capitalise]{cleveref}
\newcommand{\G}{\textbf{G }}

@@

Natural selection acts on multiple traits simultaneously. The mean trait value in a population can change either because of direct selection on trait X, or because of selection on trait Y plus a genetic correlation between X and Y \citep{lande1979quantitative,lande1983measurement}. The role of genetic correlations in the response of quantitative traits to natural selection is described by Lande's equation ($\Delta \bar{z} = \G \hspace{.05in}\beta$), where $\Delta \bar{z}$ is the expected change in the population mean of a trait $\Delta \bar{z}$, $\beta$ is the selection gradient, and the components of the \G  matrix are the additive genetic covariances between traits. Lande's equation governs the short-term evolution of correlated quantitative traits under natural selection, but it may play a role in bridging the gap between micro- and macro-evolution \citep{Arnold2001,Arnold2008,Schluter1996,Schluter2000}.\par

When the \G matrix is stable, it can be used to extrapolate future trait evolution from historical patterns, and to infer past selection gradients. The requirement for a stable \G matrix has motivated empirical work to measure and compare \G in related species. At least portions of the \G matrix may be conserved among certain phylogenetic groups \citep{Arnold1999,Roff2000,Steppan2002}, but other studies also indicate that \G is unstable. ***

Theoretical quantitative genetic models have also been used to examine the conditions under which the stability of the \G matrix can be expected \citep{Turelli1988}.  \par
% * <masel@email.arizona.edu> 2017-10-06T23:52:45.315Z:
% 
% > other studies also indicate that \G is unstable
% Citations(s) needed for empirical work for instability of G and theoretical work either deriving or assuming stability.
% 
% ^.

The genetic correlations described by \G are due to some combination of pleiotropy and linkage disequilibrium \citep{Saltz2017}. Historically, pleiotropy was favored by the Edinburgh school, who argued that recombination would quickly eliminate linkage disequilibrium \citep[Chapter~20]{fox2006evolutionary}, particularly in the randomly mating animal populations that the Edinburgh school focused on. In contrast, quantitative geneticists in the alternative, Birmingham school largely worked on inbred lines of plants, where the effects of linkage disequilibrium were impossible to ignore. The Edinburgh view came to be widely accepted, and as a result, most models for the evolution of the \G  matrix assume that genetic correlations are pleiotropic and go on to examine how evolutionary processes influence them \citep{Turelli1988}. We know far less about the about the dynamics of genetic correlations arising from linkage disequilibrium.\par
% * <masel@email.arizona.edu> 2017-10-24T22:58:18.095Z:
% 
% > most models for the evolution of the \G  matrix assume that genetic correlations are pleiotropic and go on to examine how evolutionary processes influence them \citep{Turelli1988
% More citations needed
% 
% ^.

Biology today has come to recognize the dominance of relatively asexual microbes \cite{mcfall2013animals}, for whom the Birmingham emphasis on linkage disequilibrium as a source of genetic correlations cannot be ignored. Linkage disequilibria certainly persist long enough to affect adaptation in asexuals; this fact has, for example, been the basis of explanations for the evolutionary advantages of sex and recombination \citep{Barton2005,Otto2009}. \par
% * <masel@email.arizona.edu> 2017-10-06T22:23:25.138Z:
% 
% > Biology today has come to recognize the dominance of relatively asexual microbes. 
% Cite www.pnas.org/cgi/doi/10.1073/pnas.1218525110 and maybe others making complementary points.
% 
% ^ <kgomez81@math.arizona.edu> 2017-12-05T18:40:59.153Z:
% 
% I added the citation.
%
% ^.
Asexual evolution is normally approached using a population genetic rather than a quantitative genetic approach. In particular, traveling wave models have been successfully used to describe the rate of multi-locus adaptation in asexuals given clonal interference \citep{rouzine2003solitary,desai2007beneficial}.  The basic model has only three parameters: fixed population size $N$, beneficial mutation rate $U$, and per-beneficial-mutation selection coefficient $s$. Adaptation, i.e. the increase in mean population fitness, occurs at the rate given by Desai and Fisher \citep[Equation 41]{desai2007beneficial}, expressed as
% * <masel@email.arizona.edu> 2017-10-14T17:09:10.651Z:
% 
% > Adaptation, i.e. the increase in mean population fitness, occurs at rate
% Cite the specific equation number in Desai & Fisher. Also give the parameter value conditions for which this equation holds.
% 
% ^ <kgomez81@math.arizona.edu> 2017-12-06T20:47:24.968Z:
% 
% I added the equation number to the citation.
%
% ^.
\begin{equation}\label{eq:1}
v(U,N,s) = s^2\frac{2\log(N s)-\log(s/U)}{\log^2(s/U)}.
\end{equation}
The expression holds when the asexual population is adapting in what is defined by Desai and Fisher as the ``concurrent mutations regime.'' In a monoclonal population, selection will cause a beneficial mutation to spread until all individuals carry the new mutation, at which point, it is said to have fixed in the population. The time required for beneficial mutation to fix is called its fixation time. For asexuals adapting in the concurrent mutations regime, beneficial mutations appear in rapid succession relative to the fixation time. This produces many mutant lineages that vary in the numbers of beneficial mutations that have accumulated in their genome. Beneficial mutations that appear on less fit lineages are lost due to clonal interference. Increasing the number of beneficial mutations that enter the population increases the proportion of beneficial mutations that are wasted.  The logarithmic dependence of $v(U,N,s)$ on the parameters captures these effects. An asexual population is in the concurrent mutations regime when the parameters satisfy the inequality $NU \gtrsim 1/\ln(Ns)$.  \par

Equation \ref{eq:1}, combined with Fisher's fundamental theorem, also gives the additive variance in fitness $\sigma^2$. With only one adaptive trait,  $\sigma^2$ also gives that trait's additive variance. With two adaptive traits, Equation \ref{eq:1} still gives overall fitness variance $\sigma^2$, but nothing can be said about the variances and covariance of the two traits.\par

Adaptation in one trait proceeds more slowly when there is clonal interference with mutations that are adaptive in a second trait. Quantitatively, when the two adaptive traits each have mutations with fitness benefit $s$ appearing at rate $U$, the overall adaptation rate is $v(2U,N,s)$, and so by symmetry, adaptation in each trait alone occurs at rate $0.5 \hspace{.02 in} v(2U,N,s)$. This is lower than the rate $v(U,N,s)$ when the other trait is not also evolving. The reduction in trait-specific adaptation rate must be distributed between the two terms of Lande's equation $v_1 =\sigma_1^2 +\sigma_{1,2}$. Here, we will calculate how the reduction in trait-specific adaptation partitions between the variance in that trait and its covariance with the other.\par

% * <masel@email.arizona.edu> 2017-07-29T13:00:14.234Z:
% 
% > Clearly, we lack any information about how trait two is affecting trait one's evolution,
% Spell out what exactly it is that we don't know, ie the degree to which the reduction in adaptation rate of trait 1 is due to a change in variance or a change in covariance.
% 
% 
% ^ <kgomez81@math.arizona.edu> 2017-08-22T14:11:45.747Z.

To explore how adaptation in one trait is affected by evolution in another, we extend the traveling wave model approach to a two dimensional trait space. We assume that each of the two fitness-determining traits is controlled by different loci, i.e. there is no pleiotropy. With pleiotropic contributions to \G excluded by design , we use this model to examine the dynamics of genetic correlations arising solely from linkage disequilibrium.

@@

We consider an asexual population of haploids with fixed total population size $N$. To keep the model simple, we assume that beneficial mutations occur with the same rate of $U$ mutations per birth per trait with fitness benefit $s$ in each trait and no epistasis. Deleterious mutations are not considered. Individuals with $i$ mutations in the first trait and $j$ in second, have relative fitness $r_{i,j}= i s+j s$ and absolute fitness $r_{i,j}-\bar{r}$, where $\bar{r}$ is the population mean fitness. We refer to each such set of individuals as a ``class'' and denote their abundances and frequencies by $n_{i,j}$ and $p_{i,j}$. \par

We assume $1/N \ll s \ll 1$.  Following the convention of Desai and Fisher, we refer to the set of classes abundant enough to behave deterministically $(n_{i,j}>1/s)$ as the bulk, and to smaller adaptive classes ($r_{i,j}>\bar{r}$) as the stochastic front (Figure \ref{fig:1}, solid and patterned, respectively). The stochastic front shares the same role of the nose in Desai and Fisher's traveling wave model \cite{desai2007beneficial}. Like the nose, the stochastic front contains the beneficial mutations (arrows in Figure \ref{fig:1}) that will drive future adaptation in the bulk; mutations within the bulk are wasted. However, the front in our model is evolving boundary rather than a point. Classes in the front become part of the bulk once they have ``established'' (grown sufficiently large). This shifts the position of  the bulk, with the stochastic front shifting also, to form a two dimensional traveling wave. \par 
% * <masel@email.arizona.edu> 2017-08-22T19:17:50.152Z:
% 
% > those that grow stochastically and have a positive selective advantage ($r_{i,j}>\bar{r}$) will be referred to as the stochastic front
% This is a good place to point a key difference between you and prior work, namely that your front is a curve not a point.
% 
% ^ <kgomez81@math.arizona.edu> 2017-08-23T17:03:48.195Z:
% 
% I added a bit more after this sentence, and changed the subsequent one one more sentence to indicate the differences in between Desai and Fisher's traveling wave nose. "The stochastic front shares the same role of the nose in Desai and Fisher's traveling wave model\cite{desai2007beneficial}. Like the nose, the stochastic front contains the beneficial mutations (arrows in Figure \ref{fig:1}) that will drive future adaptation in the bulk; mutations within the bulk are wasted."
% 
% ^.

\begin{figure}[!ht]
\includegraphics[width=1\linewidth]{figures/fig1a.pdf}
\caption{Illustration of a two dimensional trait distribution taken from a simulation. Individuals having an equal number of beneficial mutations in each trait are grouped into classes (boxes). Classes growing deterministically form part of the bulk (solid interiors), and those growing stochastically belong to stochastic front (patterned interiors). The red dot marks the average number of beneficial mutations in each trait. Fitness-isoclines identify the set of equally fit classes and are parallel the population mean fitness isocline (dashed red line). In asexuals, beneficial mutations must occur on the fittest backgrounds in order to contribute to the adaptive process. These mutations are represented by arrows from the bulk into the stochastic front. Over time, classes in stochastic front become part of the bulk once their abundances have become sufficiently large, while new ones subsequently emerge and become part of the stochastic front. Classes in the bulk below the population mean fitness isocline dwindle in size by selection. The distribution shifts direction of increasing relative fitness as result of these two processes and produce a two-dimensional traveling wave. (Simulation parameters: $N=10^9$, $s=0.02$, and $U=10^{-5}$)}\label{fig:1}
% * <masel@email.arizona.edu> 2017-10-14T18:05:49.152Z:
% 
% > These groups are represented by the lines passing through their centers that are parallel to the dashed line.
% Do you want to put these lines into the figure, or take this sentence out of the figure legend?
% 
% 
% ^.
\end{figure}

Changes in abundances within the bulk follow a system of ordinary differential equations given in \cite[page 27]{Crow1970} :
\begin{equation} \label{eq:2}
\dot{n}_{i,j}(t) =\left (r_{i,j}-\frac{N(t)}{N}\bar{r}\right) n_{i,j}(t),
\end{equation}
where time $t$ is measured in generations. At capacity, Equation \ref{eq:2} gives the selective advantage of a class relative to population mean fitness, denoted $s_{i,j}$, whose value is approximately  $s_{i,j} \approx (r_{i,j}-\bar{r})$. The stochastic front contains so little of the total population that variances and covariance are effectively determined only by the frequencies of classes within the bulk of two dimensional distribution. Each iteration began with solving (using NDSolve in Mathematica) for the time-dependent abundances in the bulk governed by \ref{eq:2} over a time interval of one thousand generations.\par

Growth of classes in the front can be modeled by branching processes fed by mutants from the adjacent (and exponentially growing) classes in the bulk. These mutations are shown by black arrows in Figure \ref{fig:1}. We sample the next mutation at the front by sampling random variable $x$ from a standardized exponential and solving $ x=\int_0^{\tau} b_{i,j} n_{i,j}(t)U dt$ for $\tau$. The birth rate in our model is set to $b_{i,j}=1+s_{i,j}$ per individual per generation.

We then sampled whether the mutation would establish, by comparing a uniform random variable to the probability of fixation from Equation 16 of Desai and Fisher:

\begin{equation} \label{eq:3}
\pi_{fix} = \frac{s_{i,j}}{ 1+s_{i,j}}.
\end{equation}
After establishment of a mutant lineage at abundance $1/s_{i,j}$, it is part of the bulk, and subsequent growth is deterministic and exponential. \par

We update our system based on the first mutation to appear that is destined for establishment. The timing of establishment events is also stochastic, with later deterministic growth described by $n_{i,j}(t) = \frac{1}{s_{i,j}} \exp[{s_{i,j}(t-\tau)}]$ with the random variable $\tau$ measuring the time of establishment. We consider the related random variable $\nu = \frac{1}{s_{i,j}}\exp[-s_{i,j}\tau]$ representing the initial size at the time of first mutation that would produce the deterministic trajectory observed after establishment. Using the expression given by Desai and Fisher for the PDF of $\tau$ \citep[Equation 12]{desai2007beneficial}, we can arrive at the cumulative distribution of $\nu$:
% * <masel@email.arizona.edu> 2017-10-23T22:46:00.814Z:
% 
% > Using the expression given by Desai and Fisher for the PDF of $\tau$
% Give an Equation number
% 
% ^ <kgomez81@math.arizona.edu> 2017-12-05T19:58:30.000Z:
% 
% Added the citation for the PDF of \tau needed to get the CDF of \nu, and removed the citation to Uecker and Hermisson previously included.
% 
% ^.

\begin{equation} \label{eq:4}
P(\nu \leq \nu_0) = 1- e^{-\pi_{fix} \nu_0}
\end{equation}and used it to set the initial abundance of  the establishing class equal to $\nu$. We update the other abundances of the bulk at this time. Any classes with less than one individual were removed. The next iteration began with solving for the timecourse of the new bulk, now including the newly established class, followed by the subsequent steps outlined above.\par

\begin{figure*}[!ht]
\centering
\includegraphics[width=1\linewidth]{figures/fig2.pdf} 
\caption{Measured averages of variance (open diamonds) and covariance (closed circles) for trait one obtained from simulations, along with the theoretically predicted drop in its rate of adaptation (solid line). In each graph, two population parameters are kept constant while the third is varied to determine its effects on the mean variance and covariance of trait one induced the by evolution with the second trait.  In every case, the variance of trait one increases ($\sigma_1^2 >1$), while trait interactions result in negative covariance whose magnitude is larger than the change in variance ($|\sigma_{1,2}| > \sigma_1^2-1$), indicating that the reduction in the rate of adaptation of trait one is the result of large negative covariances.  Mean variance and covariance are insensitive to changes in the population size $N$ (pane c), but exhibit some dependence on the values of $U$ and $s$ (panes a,b). The values  $s=0.02$, $U=10^{-5}$ and $N=10^9$ where used for parameters that were kept constant.}\label{fig:2}
\end{figure*}

Each class in the stochastic front may be a mixture of lineages, each arising from a different mutation event. However, we assume that the arrival time of the first mutant lineage destined to establish is the one determines which class will join the bulk next. With some probability, the  \par
% * <masel@email.arizona.edu> 2017-10-23T22:52:29.988Z:
% 
% > Each class in the stochastic front may be a mixture of lineages, each arising from a different mutation event. However, it is the arrival time of the first mutant lineage destined to establish that determines which class will join the bulk next. \par
% Requires discussion. Errors introduced when the first establishing is earlier than the establishing time of the first arrival destined to establish.
% 
% ^.

Simulations were initialized with monoclonal populations at carrying capacity consisting of individuals having no beneficial mutations in either trait and relative fitness set to zero. We varied either $N$, $s$ and $U$ while fixing the others. Each was varied across a range constructed from values reported by evolutionary experiments involving either E. coli or S. cerevisiae \citep{desai2007speed,Levy2015,Perfeito2007}.  \par
% * <masel@email.arizona.edu> 2017-10-23T22:58:03.572Z:
% 
% > We varied either $N$, $s$ and $U$ while fixing the others. Each was varied across a range constructed from values reported by evolutionary experiments involving either E. coli or S. cerevisiae
% More detail on what exactly we learned from each paper, and also how we kept within concurrent-mutations regime, including saying what the concurrent mutations regime, or do the latter around Equation 1, and refer to it here.
% 
% ^ <kgomez81@math.arizona.edu> 2017-12-06T20:49:15.515Z:
% 
% I added a small discussion of the concurrent mutations regime near equation 1.
%
% ^.

The evolution of the population can be broken down into the evolution of the two traits, according to Lande's equation:
\begin{equation}\label{eq:5}
\left[
\begin{array}{c}
\dot{\bar{r_1}} \\
\dot{\bar{r_2}} 
\end{array}
\right]
=
\left[
\begin{array}{cc}
\sigma_1^2 & \sigma_{1,2} \\
\sigma_{1,2} & \sigma_2^2 
\end{array}
\right]
\left[
\begin{array}{c}
1 \\
1 
\end{array}
\right]
\end{equation}where  $\bar{r_1}=\sum_{i,j} p_{i,j} is_1$, $\bar{r_2}=\sum_{i,j} p_{i,j} j s_2$, $\sigma_1^2 = \sum_{i,j} p_{i,j} (r_i-\bar{r_1})^2$, $\sigma_2^2 = \sum_{i,j} p_{i,j} (r_j-\bar{r_2})^2$, and  $\sigma_{1,2} =\sum_{i,j} p_{i,j} (r_i-\bar{r_1})(r_j-\bar{r_2})$. Lande's equation above is expressed as the change in the population's mean trait values over time intervals of size $s$. \par
% * <masel@email.arizona.edu> 2017-10-21T16:44:31.244Z:
% 
% > have been normalized
% How have they been normalized?
% 
% ^ <kgomez81@math.arizona.edu> 2017-12-05T20:31:15.198Z:
% 
% Sorry, its time that's scaled (not normalized) to Lande's expression above.
% 
% ^.

We allowed the simulation to run for 5,000 generations to allow the population to achieve beneficial mutation-selection balance before collecting data. This ensured that the transient effects from the initializing would not distort the statistics collected for the distribution. Following the burn-in period of the simulation, we recorded the existing classes in the bulk and their respective abundances. The data was used to compute the marginal mean fitness and variances of each trait, as well as their covariance. From these quantities we also computed the instantaneous rate of adaptation (Equation \ref{eq:5}) for each trait and overall.\par
% * <masel@email.arizona.edu> 2017-10-23T23:12:23.801Z:
% 
% >  5,000 generation
% Need to define/explain what a generation is in the model.
% 
% ^.

>>>>

\documentclass[9pt,twocolumn,twoside]{gsajnl}
% Use the documentclass option 'lineno' to view line numbers
\usepackage{caption}
\usepackage{subcaption}
\usepackage[nameinlink,capitalise]{cleveref}
\graphicspath{{figures/}}
\newcommand{\G}{\textbf{G }}

@@

Natural selection acts on multiple traits simultaneously. The mean trait value in a population can change either because of direct selection on trait X, or because of selection on trait Y plus a genetic correlation between X and Y \citep{lande1979quantitative,lande1983measurement}. The role of genetic correlations in the response of quantitative traits to natural selection is described by Lande's equation ($\Delta \bar{z} = \G \hspace{.05in}\beta$), where $\Delta \bar{z}$ is the expected change in the population mean of a trait $\Delta \bar{z}$, $\beta$ is the selection gradient, and the components of the \G  matrix are the additive genetic covariances between traits. Lande's equation governs the short-term evolution of correlated quantitative traits under natural selection. Genetic variances and covariances can also evolve over appreciable time scales, and as a result, Lande's expression alone cannot tell us how correlated quantitative traits will change in the long-run.

When the \G matrix is stable, it is possible to infer past selection gradients and extrapolate future trait evolution (both direction and rate) from historical patterns. For example, studies of past macro-evolution in populations and species have used net selection gradients to draw connections between morphological evolution and reproductive isolation \citep{price1984genetic}, infer morphological and phylogenetic relations \citep{schluter1984morphological}, and determine the extent of local adaptation between populations \citep{dudley1996}. Similar work has looked the impact of genetic constraints on rates of adaptation \citep{teplitsky2014assessing,teplitsky2011quantitative}, their ability to bias phenotypic evolution over millions of years \cite{Schluter1996}.

The importance of \G's stability in evolutionary inferences such as these has motivated a large body of empirical work focused on measuring and comparing it between populations. Comparative studies over the decades have used different metrics and statistical procedures to check whether populations have similar \G matrices. Each of these metrics vary in their ability to detect conserved structure in genetic variances and covariances \cite{Steppan2002}. Matrix correlations and maximum liklihood tests, for example, often test for correlations between \G matrices or and their components, but do not give additional information about other important characteristics that can be shared between \G matrices. Methods based on the use of principle components, such as common principle component analysis (CPCA), in contrast, can compare features between \G matrices such as shape, size, and orientation. These methods have gained popularity for this reason, and because they are capable of revealing which evolutionary forces may have significantly influenced the past dynamics of the population's \G matrix \cite{Phillips1999,Arnold2008}.

A significant number of comparative studies examining either natural populations or laboratory experiments show that at least some features of the \G matrix are relatively stable. In over 70\% of comparisons between conspecific groups and species find that \G matrices have similar shape and orientation \cite{Arnold2008}. These include populations that have diverged over millions of years \cite{Arnold1999}, and across distinct environments \cite{puentes2016similarity,teplitsky2014assessing}, suggesting that aspects of genetic constraints can be maintained over significant time scales and across a diverse set of environmental factors. There are also comparative studies that find no related structure between \G matrices \cite{wilkinson1990resistance,roff1999does,Steppan1997,Waldmann2000}, and others that reveal large instabilities in \G over very short time scales  \cite{bjorklund2013rapid,Siren2017}. 
% \cite{roff1999does} \cite{Begin2003,Begin2004}

Quantitative genetic models of genetic variances and covariances provide some indication of how ecological and evolutionary forces contribute to stability in \G's dynamics \citep{Turelli1988,barton1989evolutionary,reeve2000predicting}. For example, changes in genotype $\times$ environment interactions directly affect genetic variances and covariances and lead to significant changes in \G over few generations. Other critical stability conditions involve the strength of selection and mutational covariances. Further analytical insights from these models require knowledge of unknown genetic details, such as the distribution of allelic effects \cite{burger2000mathematical}. Simulations have been used to bypass this problem and examine the dynamics of \G in more detail \cite{Jones2003,Jones2007}. In both theoretical analysis and numerical simulations, selection and mutations play an important role in ensuring the stability of the \G matrix, but the effects of the latter are largely assumed by most quantitative geneticists to be the result of pleiotropic effects \cite{lande1979quantitative,lande1980genetic,Arnold2008}(I have more citations, and will insert them).

The genetic correlations described by \G are due to some combination of pleiotropy and linkage disequilibrium \cite{Saltz2017}. Historically, pleiotropy was favored by the Edinburgh school, who argued that recombination would quickly eliminate linkage disequilibrium \citep[Chapter~20]{fox2006evolutionary}, particularly in the randomly mating animal populations that the Edinburgh school focused on. In contrast, quantitative geneticists in the alternative, Birmingham school largely worked on inbred lines of plants, where the effects of linkage disequilibrium were impossible to ignore. The Edinburgh view came to be widely accepted, and as a result, most models for the evolution of the \G  matrix assume that genetic correlations are pleiotropic and go on to examine how evolutionary processes influence them \citep{Turelli1988}. We know far less about the about the dynamics of genetic correlations arising from linkage disequilibrium.\par

To study linkage disequilibrium, we have to consider explicit alleles or genotypes, not just quantitative traits. Beneficial mutations are discrete not infinitesimal, and appear on distinct genetic backgrounds. Selective sweeps compete with one another, and so many beneficial mutations are lost in a process known as ``clonal interference''. This will slow down adaptation \citep{hill1966effect} unless recombination brings them together, reducing the negative linkage disequilibrium between competing beneficial mutations \cite{fisher1930genetical,muller1932some}. Classical models of clonal interference considered only two loci each with two alleles. More recently, traveling wave models have been developed to describe clonal interference among large numbers of loci, arising at an arbitrary mutation rate \citep{rouzine2003solitary,desai2007beneficial,park2010speed,good2012distribution,neher2010rate,fisher2013asexual,rouzine2010multi,rouzine2007highly,rouzine2008traveling}. These models have documented that clonal interference can have an enormous impact on adaptation rates. However, to date, most of these models have treated only the evolution of ``fitness'' rather than the evolution of individual traits and their correlations \cite{}. \par

% * <masel@email.arizona.edu> 2018-01-24T22:40:30.992Z:
% 
% > owever, to date, these models have treated only the evolution of ``fitness'' rather than the evolution of individual traits and their correlations. 
% You mentioned one paper that might be an exception - find it!
% 
% ^ <kgomez81@math.arizona.edu> 2018-02-10T00:01:15.252Z:
% 
% I am still looking for this reference.
%
% ^.
% * <masel@email.arizona.edu> 2018-01-24T22:37:04.241Z:
% 
% >  \citep{rouzine2003solitary,desai2007beneficial,park2010speed,good2012distribution}
% You can include sexual models here too, eg by Neher
% 
% ^ <kgomez81@math.arizona.edu> 2018-02-09T23:38:56.909Z:
% 
% I added the Neher paper, and other references. I'll conduct another search for other sources. 
% 
% ^.

Here, we focus on asexual populations, because they have the strongest clonal interference and resulting linkage disequilibrium. Moreover, biology today has come to recognize the numerical prevalence and ubiquitous significance of relatively asexual microbes \cite{mcfall2013animals}.\par

We begin with Desai and Fisher's \citep{desai2007beneficial} framework of fixed population size $N$, beneficial mutations at rate $U$, and per-beneficial-mutation selection coefficient $s$. This yields the rate of adaptation \citep[Equation 41]{desai2007beneficial}
\begin{equation}\label{eq:1}
v(U,N,s) = s^2\frac{2\log(N s)-\log(s/U)}{\log^2(s/U)}.
\end{equation}
Equation \ref{eq:1} holds when the parameters satisfy $NU \gtrsim 1/\ln(Ns)$ and $U/s \ll 1$ (the ``concurrent mutations regime''  where beneficial mutations appear rapidly relative to the time required for any one of them to fix). When combined with Fisher's fundamental theorem, Equation \eqref{eq:1} also gives the additive variance in fitness $\sigma^2$. When there is only one adaptive trait, $\sigma^2$ is that trait's additive variance. \par

With two adaptive traits, Equation \eqref{eq:1} still provides overall fitness variance $\sigma^2$, but it does not give its decomposition into the variances and covariance expressed in Lande's equation. Consider just two traits, each experiencing beneficial mutations at rate $U$ which each increase fitness by $s$. Equation \ref{eq:1} indicates that the overall adaptation rate is $v(2U,N,s)$, and so by symmetry, adaptation in each trait alone occurs at rate $0.5 \hspace{.02 in} v(2U,N,s)$. This is lower than the rate $v(U,N,s)$ when the other trait is not evolving. The reduction in trait-specific adaptation rate must be distributed between the two terms of Lande's equation $v_1 =\sigma_1^2 +\sigma_{1,2}$. More importantly, the contributions of each term will over time change if the \G matrix is not constant.\par

Here we analyze a two-dimensional traveling wave model of an asexual population, and find that clonal interference alone, in the absence of pleiotropy, leads on average to greatly elevated variance of fitness-associated traits and to negative covariance between them.  However, the \G matrix is highly unstable over the timescale of selective sweeps.

@@

We consider an asexual population of haploids with fixed total population size $N$ evolving in continuous time. To keep the model simple, we assume that beneficial mutations occur with the same rate of $U$ mutations per trait per birth giving a fitness benefit $s$ with no epistasis. Deleterious mutations are not considered. Since there is no epistasis, all individual with $i$ mutations in the first trait and $j$ in the second have the same Malthusian relative fitness $r_{i,j}= i s+j s$. We refer to each such set of individuals with the same $i$ and $j$ as a ``class'' and denote their abundances and frequencies by $n_{i,j}$ and $p_{i,j}$ respectively. The population's mean fitness is denoted $\bar{r}$.

We assume that $N$ is so large that selection is strong relative to drift ($1/N \ll s$). In this case, most classes are so large that their frequency grows or declines approximately deterministically due to selection. We refer to these classes as the ``bulk''  (Figure \ref{fig:1}, solid cells). However, adaptive mutant lineages arising along the high-fitness ``front'' of the population will start with a single individual, and initially behave stochastically (Figure \ref{fig:1}, patterned cells; arrows show mutations). Many of these lineages will go extinct before attaining appreciable frequencies, but some will make it to high enough abundances that they begin to grow deterministically. This transition from stochastic to deterministic behavior is called ``establishment'' and occurs if the lineage reaches an abundance of order $1/s$.  \par

The evolution of population follows an iterative process with two steps. First, the next class to establish is determined, and its depends on the growth of classes producing adaptive mutants in the front (\ref{fig:1}, arrows). Once a class has established, it begins to alter the growth dynamics of the other classes in the bulk. The second step involves producing the solutions for the dynamics of the bulk with the newly established class. Having done so, the process then repeats with determining the next class from the front to establish using the solutions obtained.

To simulate the timing of the next establishment, stochastic fluctuations induced by drift and mutations must be accounted for in the growth of classes at the front. These growth of classes in the front follow a branching processes fed by mutants from the adjacent (and exponentially growing) classes in the bulk.  Each of the mutant lineages that arise will either go extinct, or grow sufficiently large to experience deterministic growth from selection. The chances of the latter happening is given by the probability of establishment, denoted $\pi$. We use the value of $\pi$ given by the expression:
\begin{equation} \label{eq:2}
\pi = \frac{s_{i,j}}{ 1+s_{i,j}},
\end{equation}and derived from \cite[Equation 16]{desai2007beneficial}. Here $s_{i,j}=(r_{i,j}-\bar{r})$ is the selective advantage of an individual in the class with $i$ mutation in the first trait, and $j$ mutations in the second, with respect to an average individual in the population.  Although classes in the front can consist of many mutant lineages that contribute to their size during the stochastic growth phase, their establishment  is driven by the size of the first lineage destined to establish. In particular, the time required for this single lineage to transition into the deterministic growth phase is a good approximation for the actual time of establishment of its class.  The justification for this approximation is discussed in detail by \cite[p.1773]{desai2007beneficial}, and intuitively has to do with the fact that lineages from early mutation events often quickly die out. \par

The next establishment event and its corresponding class  is determined by sampling the each of the sojourn times of mutations in the front that are destined to establish and identifying the first to appear. Beneficial mutations within the bulk are ignored because these are lost to clonal interference. At the front, the number of beneficial mutations from a class $(i,j)$ in the bulk follows an inhomogeneous Poisson  process with a rate $\lambda_{i,j}(t)$. The rate will depend on the birth rate of individuals in the class $(i,j)$, their time-dependent abundances, and the beneficial mutation rate (see appendix).  The sojourn time $\tau$ for the appearance of the first mutation destined to establish can be transformed into the random variable $x=\int_0^\tau \pi_{fix} \lambda_{i,j}(t) dt$ that has a standard exponential distribution. The sampling of $\tau$ is be done by generating a value for  $x$ and numerically solving for $\tau$ in the integral equation. This is repeated for each class of the bulk adjacent to the stochastic front, and the next class to establish is picked to be the one with the smallest sojourn time. \par

A class that is chosen to establish is immediately incorporated into the bulk with an initial size that reflects stochastic effects on the growth of the single mutant lineage that drives its establishment. If a mutant lineage that is destined to establish appears at time $t=0$, it requires $T$ generations to grow to size $1/s_{i,j}$. The variable $T$ is random due to the effects of drift on the growth of the lineage, but its distribution is known and given by \cite[Equation 12]{desai2007beneficial}.  Once this lineage establishes $T$ generations later, the time dependent growth of its class is approximately $n_{i,j}(t) =s_{i,j}^{-1} \exp[{s_{i,j}(t-T)}]$. We consider instead an initial size  $\nu = s_{i,j}^{-1}\exp[-s_{i,j}T]$, and equivalently regard $n_{i,j}(t) = \nu s_{i,j}^{-1} \exp[{s_{i,j}t}] $ as describing the deterministic growth of the class beginning at time $t=0$. Consequently, each class that establishes can be incorporated into the bulk from the moment at which mutation driving establishment appears, but its initial size will be $\nu$. The random variable $\nu$ can be sampled by using its cumulative distribution function:
\begin{equation}\label{eq:3}
P(\nu \leq \nu_0) = 1- e^{-\pi_{fix} \nu_0},
\end{equation}
which is obtained from the probability density function of  $T$ in \cite[Equation 12]{desai2007beneficial}.

Following establishment, we update the set of classes in the bulk and calculate solutions to their time-dependent growth. Initially, classes with less than one individual are regarded as having gone extinct and removed from the bulk. Abundances of the remaining class and the newly established one change at a rate of  $\dot{n}_{i,j}(t) =(r_{i,j}-\bar{r})n_{i,j}(t)$ due to selection. Initial values are the size of each class at the time of establishment. Changes in the set of frequencies $p_{i,j}$ obey the same ordinary differential equation because $N$ is constant. Solutions of the coupled system of ODE's are then obtained using an ODE solver (NDSolve in Mathematica) for a period of one thousand generations, and are then used to determine the next establishment event. The process then continues for the duration of the simulation. Note that in our implementation the population size $N(t)=\sum n_{i,j}(t)$ can deviate slightly from the value $N$ as establishments take place and classes in the bulk become extinct. To account for this and maintain a total population size as close to $N$ as possible, we instead use the ODE 
\[ \dot{n}_{i,j} = (r_{i,j}-\bar{r})n_{i,j} - \bar{r}\left (\frac{N(t)}{N}-1 \right)n_{i,j} \] 
given by \cite[p.133]{Crow1970} in our simulations.\par

Simulations were initialized with monoclonal populations at carrying capacity consisting of individuals having one beneficial mutation in either trait and relative fitness set to zero. We varied either $N$, $s$ and $U$ while fixing the others. Each was varied across a range constructed from values reported by evolutionary experiments involving either E. coli or S. cerevisiae \citep{desai2007speed,Levy2015,Perfeito2007}.  We allowed the simulation to run for 5,000 generations, where one generation  This allowed the population to achieve beneficial mutation-selection balance before collecting data, and ensured that the transient effects from the initializing would not distort the statistics collected for the distribution. Following the burn-in period of the simulation, we recorded the existing classes in the bulk and their respective abundances. \par

The data was used to compute the marginal mean fitness and variances of each trait, as well as their covariance. From these quantities we also computed the instantaneous rate of adaptation (Equation \eqref{eq:5}) for each trait and overall.
The evolution of the population can be broken down into the evolution of the two traits, according to Lande's equation:
\begin{equation}\label{eq:5}
\left( \begin{array}{c}
\dot{\bar{r_1}} \\
\dot{\bar{r_2}} 
\end{array} \right)
=
\left( \begin{array}{cc}
\sigma_1^2 & \sigma_{1,2} \\
\sigma_{1,2} & \sigma_2^2 
\end{array}\right)
\left(\begin{array}{c}
1 \\
1 
\end{array}\right),
\end{equation}
where  $\bar{r_1}=\sum_{i,j} p_{i,j} is_1$, $\bar{r_2}=\sum_{i,j} p_{i,j} j s_2$, $\sigma_1^2 = \sum_{i,j} p_{i,j} (r_i-\bar{r_1})^2$, $\sigma_2^2 = \sum_{i,j} p_{i,j} (r_j-\bar{r_2})^2$, and  $\sigma_{1,2} =\sum_{i,j} p_{i,j} (r_i-\bar{r_1})(r_j-\bar{r_2})$. Here, it is the frequencies corresponding to the bulk that are relevant, since classes in the front are assumed to be too small to affect the means, variances and covariances of the distribution  ($1/s \ll N$).   Lande's equation above is expressed as the change in the population's mean trait values over time intervals of size $s$. \par
